---
title: "cpaltemplates"
subtitle: "R Package for CPAL Data Team Workflow Standardization"
format:
  html:
    toc: false
    page-layout: full
execute:
  echo: true        # Show all code chunks
  eval: false       # Don't execute any code chunks by default
  warning: false    # Don't show warnings
  message: false    # Don't show messages
  error: false      # Don't show errors
  include: true     # Include chunks in output
---

## Welcome to cpaltemplates

The **cpaltemplates** package transforms how the Child Poverty Action Lab (CPAL) data team creates analyses, reports, and visualizations. **Instead of spending time on formatting and setup, you can focus on insights and impact.**

### **Why cpaltemplates Exists**

**The Challenge:** CPAL data team members were spending significant time on repetitive tasks:
- **30-60 minutes** setting up each new project with proper folder structure, version control, and dependencies
- **Inconsistent branding** across visualizations, reports, and presentations  
- **Manual styling** of every plot, table, and document to match CPAL standards
- **Difficult onboarding** for new team members learning CPAL's preferred workflows
- **Accessibility gaps** when visualizations weren't tested for colorblind users or proper contrast

**The Solution:** cpaltemplates automates professional best practices:
- **Quick project setup** with enterprise-grade workflows built-in
- **Automatic CPAL branding** for all visualizations, reports, and dashboards
- **Built-in accessibility** ensuring inclusive design from the start
- **Standardized workflows** that scale from simple analysis to complex multi-output projects
- **Team consistency** with shared templates, functions, and quality standards

### **Impact on Your Workflow**

| **Before cpaltemplates** | **With cpaltemplates** | **Time Saved** |
|-------------------------|----------------------|----------------|
| Manual project setup with inconsistent structures | `start_project()` with standardized templates | 45+ minutes per project |
| Custom styling each plot individually | `theme_cpal()` + `scale_color_cpal()` automatic styling | 10+ minutes per visualization |
| Manual logo placement and sizing | `add_cpal_logo()` with smart positioning | 5+ minutes per branded output |
| Learning complex interactive libraries | `cpal_interactive()` with one-function conversion | 30+ minutes per interactive viz |
| Manual accessibility checking | `check_plot_accessibility()` automated testing | 15+ minutes per visualization |
| Inconsistent table formatting | `cpal_table_gt()` professional styling | 20+ minutes per table |

**Total time savings:** **2+ hours per typical analysis project**, plus immeasurable improvements in consistency and professional quality.

## Package Scope and Capabilities

cpaltemplates provides **59 functions across 7 integrated systems** that work together seamlessly:

::: {.grid}

::: {.g-col-12 .g-col-md-6}
### 🎨 **Visual Design Foundation**

**Colors & Palettes (7 functions)**
- **23 distinct color palettes** Designed for different data types
- **Accessibility built-in**: All palettes tested for colorblind users
- **Brand consistency**: Midnight (#004855) as primary brand color, plus teal, pink, orange, gold
- **Data-appropriate**: Sequential, diverging, categorical palettes for proper data representation

```{r}
# Automatic brand-consistent colors
scale_fill_cpal(palette = "main")
view_cpal_palettes("all")  # Interactive exploration
```

**Themes & Styling (8 functions)**  
- **6 theme variants** optimized for different outputs (reports, presentations, web, print, maps, dark mode)
- **Google Fonts integration**: Inter and Roboto with cross-platform fallbacks
- **Customizable parameters**: Grid lines, axis styling, legend positioning
- **Session-wide consistency**: `set_theme_cpal()` for automatic application

```{r}
# Professional theming for any context
theme_cpal()           # Most reports and presentations
theme_cpal_dark()      # Dark backgrounds and night viewing
theme_cpal_print()     # PDFs and printed materials
```
:::

::: {.g-col-12 .g-col-md-6}
### 📊 **Data Presentation Tools**

**Plots & Visualizations (8 functions)**
- **Professional output management**: `save_cpal_plot()` with preset dimensions for different uses
- **Smart logo integration**: `add_cpal_logo()` with automatic theme detection
- **Accessibility assurance**: `check_plot_accessibility()` for inclusive design
- **Explicit scale control**: Clear continuous vs. discrete color mapping

```{r}
# Professional plot workflow
plot + theme_cpal() |> 
  add_cpal_logo() |>
  save_cpal_plot("output.png", size = "slide")
```

**Tables & GT Styling (2 functions)**
- **Static tables**: GT integration with CPAL branding and advanced formatting
- **Interactive tables**: Reactable integration with sorting, filtering, data bars
- **Consistent typography**: Automatic font coordination with plot themes
- **Multiple output formats**: Optimized for reports, dashboards, presentations

```{r}
# Professional table creation
cpal_table_gt(data, title = "Executive Summary",
              highlight_columns = c("key", "metrics"))
```
:::

:::

::: {.grid}

::: {.g-col-12 .g-col-md-6}
### ⚡ **Interactive & Dynamic Features**

**Interactive Features (9 functions)**
- **ggiraph integration**: Transform any CPAL plot into interactive exploration
- **Geographic mapping**: Mapbox integration with Dallas-area defaults
- **Font management**: Google Fonts working correctly in web contexts
- **Specialized interactivity**: Point, line, column, polygon interactive geoms

```{r}
# Interactive visualization
ggplot(data, aes(x = x, y = y)) +
  cpal_point_interactive(aes(tooltip = "details")) +
  theme_cpal() |>
  cpal_interactive()
```

### 🛠️ **Project Infrastructure**

**Project Setup (17 functions)**
- **6 project types**: Analysis, Quarto reports/slides/web, Shiny dashboards/apps, R packages
- **Modern workflows**: renv, git, GitHub, targets, testing integration
- **Template system**: Professional starting code for every project type
- **CPAL asset integration**: Logos, CSS, fonts automatically included

```{r}
# Complete project setup in minutes
start_project(
  name = "impact-analysis",
  project_type = "analysis", 
  features = c("renv", "git", "targets")
)
```
:::

::: {.g-col-12 .g-col-md-6}
### 🔧 **Enhancement & Management**

**Utilities & Helpers (12 functions)**
- **Existing project enhancement**: Add CPAL capabilities to any project
- **Asset management**: Update branding, access specific logos/CSS
- **Template addition**: Quarto reports, Shiny apps, workflow automation
- **Font utilities**: Consistent typography across all contexts

```{r}
# Enhance existing projects
use_quarto_report()     # Add professional reporting
use_shiny_dashboard()   # Add interactive dashboard
update_cpal_assets()    # Refresh branding
```

### 📋 **Team Collaboration**

**Workflow Integration**
- **Complete end-to-end examples**: 5 comprehensive workflow types
- **Team standardization**: Shared functions and quality standards
- **Automation patterns**: Reproducible analysis pipelines
- **Knowledge sharing**: Documentation and troubleshooting resources

```{r}
# Team-standard project creation
create_cpal_standard_project("analysis-name", 
                            lead_analyst = "Your Name")
```
:::

:::

## Who Should Use cpaltemplates

### **Perfect for CPAL Data Team Members Who...**

| **Your Situation** | **How cpaltemplates Helps** | **Key Benefits** |
|-------------------|---------------------------|------------------|
| **Create visualizations for reports or presentations** | Automatic CPAL branding, professional themes, accessibility testing | Consistent brand presence, time savings |
| **Start new analysis projects regularly** | Automated project setup with best practices | 45+ minutes saved per project |
| **Collaborate with other team members** | Standardized templates and workflows | Seamless collaboration and handoffs |
| **Create interactive dashboards or applications** | Pre-configured Shiny templates with CPAL theming | Professional dashboards without UI expertise |
| **Generate regular reports or presentations** | Quarto templates with CPAL branding and automation | Consistent professional outputs |
| **Need to make visualizations accessible** | Built-in accessibility testing and colorblind-safe palettes | Inclusive design without expertise |
| **Want to follow modern R best practices** | renv, git, targets integration in all project templates | Professional workflows without configuration |

### **Less Useful If You...**

- Work primarily on non-CPAL projects (package is CPAL-specific)
- Need completely custom styling that conflicts with CPAL brand standards
- Do only quick, internal exploratory analysis with no sharing or presentation needs
- Work in environments where package installation is restricted

## Quick Start Journey

### **First-Time Setup**

```{r}
# Quick Start section
library(cpaltemplates)
packageVersion("cpaltemplates")

setup_cpal_google_fonts()
view_cpal_palettes("all")  # See all color options

# First-Time Setup section  
view_cpal_palettes("all")  # Interactive exploration
```

### **Create Your First Professional Visualization**

```{r}
library(ggplot2)

# Your first CPAL-branded visualization
first_plot <- ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_cpal(palette = "main") +     # CPAL colors
  theme_cpal() +                           # CPAL theme
  labs(
    title = "Vehicle Weight vs Fuel Efficiency",
    subtitle = "Professional visualization with CPAL branding",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon", 
    color = "Cylinders",
    caption = "Data: mtcars dataset"
  ) |>
  add_cpal_logo(position = "top-right") |> # CPAL logo
  save_cpal_plot("my_first_cpal_plot.png") # Professional output

# Check accessibility
check_plot_accessibility(first_plot)
```

### **Start Your First CPAL Project**

```{r}
# Interactive setup guides you through choices
start_project()

# Or direct setup if you know what you want
start_project(
  name = "my-first-analysis", 
  project_type = "analysis",
  features = c("renv", "git", "targets")
)

# Project opens automatically with:
# ✓ Professional folder structure
# ✓ CPAL assets and branding
# ✓ Modern workflow tools (renv, git, targets)
# ✓ Template code to get started
# ✓ Documentation and next steps
```

## Real-World Impact Examples

### **Before and After: Quarterly Report Creation**

**Traditional Approach (4-6 hours):**
1. Set up project folders manually (15 min)
2. Configure Quarto with custom CSS (30 min)
3. Find and resize CPAL logos (15 min)
4. Create each visualization with manual styling (45 min each × 6 charts = 4.5 hours)
5. Format tables manually (30 min)
6. Test outputs and fix inconsistencies (45 min)

**With cpaltemplates (45 minutes):**
1. `start_project(project_type = "quarto-report")` (2 min)
2. Write analysis with automatic CPAL styling (30 min)
3. `check_plot_accessibility()` on all outputs (5 min)
4. `quarto::quarto_render()` for professional report (3 min)
5. Share with stakeholders (5 min)

**Result:** Same professional quality, **5+ hours saved**, more consistent branding, built-in accessibility.

### **Before and After: Team Onboarding**

**Traditional Approach (2-3 days):**
- Learn CPAL color codes and brand guidelines
- Set up development environment manually
- Learn project structure conventions
- Understand output formatting requirements
- Create first visualization with proper styling

**With cpaltemplates (2-3 hours):**
- `library(cpaltemplates)` and `setup_cpal_google_fonts()` 
- `start_project()` with guided setup
- Follow [Getting Started Guide](getting-started.qmd)
- Create professional visualization with automatic CPAL styling
- Access team knowledge through documentation and Notion guide

**Result:** New team members productive immediately with professional-quality outputs.

## Documentation Architecture

This documentation system is designed for **both quick reference and comprehensive learning**:

::: {.grid}

::: {.g-col-12 .g-col-md-6}
### **Function Reference Documentation**

**Core Systems:**
- [**Colors & Palettes**](colors-palettes.qmd) - 23 palettes, color theory, accessibility guidance
- [**Themes & Styling**](themes-styling.qmd) - 8 themes, typography, logo integration
- [**Plots & Visualizations**](plots-visualizations.qmd) - 8 functions, professional output management
- [**Interactive Features**](interactive-features.qmd) - 9 functions, ggiraph, mapping, dashboards

**Data Presentation:**
- [**Tables & GT Styling**](tables-gt.qmd) - 4 functions, static and interactive tables
:::

::: {.g-col-12 .g-col-md-6}
### **Workflow and Integration Guides**

**Project Management:**
- [**Project Setup**](project-setup.qmd) - 17 functions, 6 project types, modern workflows
- [**Utilities & Helpers**](utilities-helpers.qmd) - 12 functions, existing project enhancement

**Team Resources:**
- [**Workflow Integration**](workflow-integration.qmd) - Complete end-to-end workflow examples
- [**Troubleshooting**](troubleshooting.qmd) - Problem-solving guide with team support resources
:::

:::

### **Documentation Standards**

**Each documentation page includes:**
- **Why and when guidance**: Context for choosing functions and approaches
- **Real-world examples**: Complete, working code that solves actual problems
- **Decision support**: Tables, flowcharts, and matrices for choosing optimal approaches
- **Troubleshooting**: Common issues and diagnostic code for each function group
- **Best practices**: Professional standards and quality guidelines
- **Team integration**: How functions work together and with team workflows

## Getting Started Path

### **For New Team Members**

1. **Start here**: [Getting Started Guide](getting-started.qmd) - Installation, setup, first plot, project creation
2. **Explore capabilities**: Browse function group documentation to understand what's available
3. **Follow workflows**: [Workflow Integration](workflow-integration.qmd) for complete project examples
4. **Get help**: [Troubleshooting](troubleshooting.qmd) and CPAL Data Team Notion Guide for support

### **For Experienced R Users**

1. **Quick overview**: This page for package scope and capabilities
2. **Function reference**: Jump directly to relevant function group documentation
3. **Integration patterns**: [Workflow Integration](workflow-integration.qmd) for advanced usage
4. **Enhancement**: [Utilities & Helpers](utilities-helpers.qmd) for adding to existing projects

### **For Team Leaders and Project Managers**

1. **Impact assessment**: Review time savings and consistency benefits above
2. **Team adoption**: [Project Setup](project-setup.qmd) for standardized team workflows
3. **Quality assurance**: Built-in accessibility and branding consistency features
4. **Collaboration**: [Workflow Integration](workflow-integration.qmd) for multi-person project coordination

## Package Philosophy and Design Principles

### **Automation Over Configuration**

**Smart defaults that work immediately:**
- Colors automatically match CPAL brand guidelines
- Themes apply professional styling without customization
- Project templates include modern best practices by default
- Interactive features inherit CPAL styling automatically

**Customization when needed:**
- All functions provide parameters for specific adjustments
- Themes can be modified while maintaining brand consistency
- Project types can be mixed and enhanced with utility functions
- Advanced users can build on CPAL foundations

### **Accessibility First**

**Inclusive design built-in:**
- All color palettes tested for colorblind accessibility
- Automated accessibility checking with `check_plot_accessibility()`
- High contrast themes available for visibility needs
- Screen reader friendly table structures and labeling

### **Team Collaboration Focus**

**Designed for multi-person teams:**
- Consistent outputs regardless of individual styling preferences
- Shared templates and standards reduce coordination overhead
- Version control and reproducibility built into project templates
- Knowledge sharing through documentation and team resources

### **Progressive Enhancement**

**Start simple, add complexity as needed:**
- Basic functions work with minimal parameters
- Advanced features available for complex requirements  
- Existing projects can be enhanced incrementally
- New capabilities can be added without disrupting existing work

## Success Stories and Use Cases

### **Typical CPAL Use Cases**

| **Project Type** | **cpaltemplates Functions Used** | **Outcome** |
|------------------|----------------------------------|-------------|
| **Quarterly program evaluation** | Colors, themes, plots, project setup, reporting | Professional stakeholder report with consistent branding |
| **Board presentation preparation** | All visualization functions + slides + interactive dashboard | Coordinated presentation package with live data exploration |
| **Ongoing program monitoring** | Interactive features, dashboard templates, automation | Real-time monitoring system with automated alerts |
| **Research paper publication** | Full function suite with accessibility focus | Reproducible research with multiple output formats |
| **New team member onboarding** | Project templates, documentation, troubleshooting | Productive team member within hours instead of days |

### **Measurable Team Benefits**

**Quality improvements:**
- **Brand consistency** across all team outputs
- **Universal accessibility compliance** through automated testing
- **Reproducible workflows** with version control and dependency management
- **Professional presentation quality** without design expertise

## Technical Requirements and Compatibility

### **System Requirements**

**Minimum requirements:**
- R 4.0+ (required for modern package features)
- RStudio (recommended for optimal project integration)
- Internet connection (for Google Fonts and template downloads)

**Optional but recommended:**
- Git (for version control features)
- Quarto (for advanced reporting capabilities)  
- Modern web browser (for interactive features)

### **Dependency Management**

**Core dependencies** (automatically installed):
- ggplot2, cli, fs - Essential for all cpaltemplates functions

**Optional dependencies** (installed as needed):
- ggiraph, reactable, mapgl - For interactive features
- gt - For advanced table formatting
- renv, gert, yaml - For project management features
- cowplot, magick - For logo integration

**Graceful degradation:** All functions provide helpful error messages and alternatives when optional packages are missing.

## Getting Support and Contributing

### **When You Need Help**

1. **Function-specific help**: Use `?function_name` for detailed parameter documentation
2. **Common issues**: Check [Troubleshooting](troubleshooting.qmd) for systematic problem-solving
3. **Team support**: Use CPAL Data Team Notion Guide for team-specific guidance and to connect with team members
4. **Advanced workflows**: Review [Workflow Integration](workflow-integration.qmd) for complete examples

### **Contributing to Team Knowledge**

**Help improve cpaltemplates for everyone:**
- **Document solutions**: Add new troubleshooting discoveries to CPAL Data Team Notion Guide
- **Share workflows**: Contribute successful project patterns to team knowledge base
- **Report issues**: Use team communication channels to report bugs or suggest improvements
- **Mentor others**: Help onboard new team members using your cpaltemplates experience

### **Package Evolution**

**cpaltemplates continues evolving based on CPAL team needs:**
- New functions added based on common team requirements
- Templates updated to reflect changing best practices
- Branding assets refreshed to match organizational updates
- Documentation improved based on team feedback and common questions

---

## Ready to Transform Your Workflow?

**Start with what you need most:**

::: {.grid}

::: {.g-col-12 .g-col-md-4}
### **New to cpaltemplates?**
[**Getting Started →**](getting-started.qmd)

Installation, setup, and your first professional visualization in 10 minutes.
:::

::: {.g-col-12 .g-col-md-4}  
### **Need specific functions?**
[**Function Reference →**](colors-palettes.qmd)

Detailed documentation for colors, themes, plots, tables, and interactive features.
:::

::: {.g-col-12 .g-col-md-4}
### **Want complete workflows?**
[**Integration Examples →**](workflow-integration.qmd)

End-to-end workflows for analysis, reporting, dashboards, and team collaboration.
:::

:::

---

*cpaltemplates is developed and maintained by the CPAL Data Team to transform data analysis workflows from time-consuming, inconsistent processes into efficient, professional, and collaborative systems that amplify your analytical impact.*
